Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-11-18 paper-new/intro.tex
small wording change in intro

The intensional nature of guarded domain theory leads to the main departure of
our semantics from the New-Licata approach. Because casts involve computational
steps, the graduality property must be insensitive to the steps taken by terms.
This means that the model must allow for reasoning \emph{up to weak
bisimilarity}, where two terms are weakly bisimilar if they differ only in their
number of computational steps. However, the weak bisimilarity relation is not
transitive in the guarded setting, which follows from a no-go theorem we
establish (Theorem \ref{thm:no-go}). The New-Licata proofs freely use
transitivity, and we argue in Section \ref{sec:towards-relational-model} that
some amount of transitive reasoning is necessary for defining a
syntax-independent model of gradual typing. As a result, the lack of
transitivity of weak bisimilarity presents a challenge in adapting the
New-Licata model to the guarded setting. Our solution is to work with a
combination of weak bisimilarity and step-sensitive reasoning, where the latter
notion \emph{is} transitive. To deal with the fact that casts take computational
steps, we introduce the novel concept of \emph{syntactic perturbations}, which
are explicit synchronizations that we manipulate syntactically. The combination
of lock-step reasoning with perturbations is transitive and is able to account
for the step-insensitive nature of the graduality property.

>>>>

The intensional nature of guarded domain theory leads to the main departure of
our semantics from the New-Licata approach. Because casts involve computational
steps, the graduality property must be insensitive to the steps taken by terms.
This means that the model must allow for reasoning \emph{up to weak
bisimilarity}, where two terms are weakly bisimilar if they differ only in their
number of computational steps. However, the weak bisimilarity relation is not
transitive in the guarded setting, which follows from a no-go theorem we
establish (Theorem \ref{thm:no-go}). The New-Licata proofs freely use
transitivity, and we argue in Section \ref{sec:towards-relational-model} that
some amount of transitive reasoning is necessary for defining a
syntax-independent model of gradual typing. As a result, the lack of
transitivity of weak bisimilarity presents a challenge in adapting the
New-Licata model to the guarded setting. Our solution is to work with a
combination of weak bisimilarity and step-sensitive reasoning, where the latter
notion \emph{is} transitive. To deal with the fact that casts take computational
steps, we introduce the novel concept of \emph{syntactic perturbations}, which
are explicit synchronizations that we manipulate syntactically. The combination
of step-sensitive reasoning with perturbations is transitive and is able to account
for the step-insensitive nature of the graduality property.
